Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm: A Case Study

### 2022-05-04 Paper_Draft2/Draft2.Rmd
delete citation
fixing typos adding wordcount

Our project builds on Women at the Table’s various debiasing algorithms used in its AI & Equality Human Rights Toolbox to conduct our own analyses on the COMPAS data set. Based on this analysis, we employ a human rights framework to contribute to the ProPublica and Northpointe debate and investigate whether or to what extent there is racial bias in the COMPAS algorithm. With a solid understanding of the two sides, we aim to pinpoint the shortcomings of both arguments and correct them in our analyses. We will use various debiasing techniques and fairness metrics to evaluate the level of bias present in the COMPAS data and our algorithm. We will summarize our results using the JupyterNotebook framework from Women at the Table, to be used by members of the organization to teach in a workshop setting. We hope that our findings will highlight the importance of checking statistical analyses using varied methods and contribute to the ongoing discussion of the effects of machine biases in the justice system.

@@

The data we are using for this library addition is the COMPAS General Recidivism Risk Scores data set from the [AI Fairness 360](https://aif360.readthedocs.io/en/stable/modules/generated/aif360.datasets.CompasDataset.html#aif360.datasets.CompasDataset) (AIF360) toolkit. The AIF360 toolkit builds on the dataset released by ProPublica collaborators Larson et al. created to examine the racial bias and the true outcomes of the recidivism risk scores in the COMPAS algorithm for the initial ["Machine Bias"](https://www.propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing) article. For this, Larson et al. obtained two years worth of COMPAS scores from the Broward County Sheriff’s Office in Florida, as well as the corresponding intake information for each defendant including but not limited to name, sex, race, age, and charge degree and description. They also obtained data about whether defendants actually recidivated or not in the two year period following their initial COMPAS score assessment. AIF360 then processed the data with the same procedures that Larson et al. followed for their analysis.

@@

```{r table snip, echo = FALSE, fig.align='center', fig.cap= "A snippet of the data set we will be using, containing information on a defendant's age, sex, race, criminal history, charge degree, charge description, and two-year recidivism outcome.", out.width= "100%"}
knitr::include_graphics("../images/table_snippet.png")
```

@@

AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks “to help facilitate the transition of fairness research algorithms to use in an industrial setting and to provide a common framework for fairness researchers to share and evaluate algorithms” [@aif360-oct-2018]. It contains multiple data sets, including the COMPAS data set that accompanied @angwin2016machine.

@@

This metric is best used when it is very important to catch positive outcomes while false positives are not exceptionally problematic [@Cortez2019How]. This is not the case for the COMPAS data set, as false positives mean extra jail time for someone who will not actually re-offend.

@@

>>>>

Our project builds on Women at the Table’s various debiasing algorithms used in its AI & Equality Human Rights Toolbox to conduct our own analyses on the COMPAS dataset. Based on this analysis, we employ a human rights framework to contribute to the ProPublica and Northpointe debate and investigate whether or to what extent there is racial bias in the COMPAS algorithm. With a solid understanding of the two sides, we aim to pinpoint the shortcomings of both arguments and correct them in our analyses. We will use various debiasing techniques and fairness metrics to evaluate the level of bias present in the COMPAS data and our algorithm. We will summarize our results using the JupyterNotebook framework from Women at the Table, to be used by members of the organization to teach in a workshop setting. We hope that our findings will highlight the importance of checking statistical analyses using varied methods and contribute to the ongoing discussion of the effects of machine biases in the justice system.

@@

The data we are using for this library addition is the COMPAS General Recidivism Risk Scores dataset from the [AI Fairness 360](https://aif360.readthedocs.io/en/stable/modules/generated/aif360.datasets.CompasDataset.html#aif360.datasets.CompasDataset) (AIF360) toolkit. The AIF360 toolkit builds on the dataset released by ProPublica collaborators Larson et al. created to examine the racial bias and the true outcomes of the recidivism risk scores in the COMPAS algorithm for the initial ["Machine Bias"](https://www.propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing) article. For this, Larson et al. obtained two years worth of COMPAS scores from the Broward County Sheriff’s Office in Florida, as well as the corresponding intake information for each defendant including but not limited to name, sex, race, age, and charge degree and description. They also obtained data about whether defendants actually recidivated or not in the two year period following their initial COMPAS score assessment. AIF360 then processed the data with the same procedures that Larson et al. followed for their analysis.

@@

```{r table snip, echo = FALSE, fig.align='center', fig.cap= "A snippet of the dataset we will be using, containing information on a defendant's age, sex, race, criminal history, charge degree, charge description, and two-year recidivism outcome.", out.width= "100%"}
knitr::include_graphics("../images/table_snippet.png")
```

@@

AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks “to help facilitate the transition of fairness research algorithms to use in an industrial setting and to provide a common framework for fairness researchers to share and evaluate algorithms” [@aif360-oct-2018]. It contains multiple datasets, including the COMPAS dataset that accompanied @angwin2016machine.

@@

This metric is best used when it is very important to catch positive outcomes while false positives are not exceptionally problematic [@Cortez2019How]. This is not the case for the COMPAS dataset, as false positives mean extra jail time for someone who will not actually re-offend.

@@

This document contains 4793 words
